Liveness-Based Garbage Collection for Lazy Languages

### 2015-03-21 paper/paper.tex
End of liveness

\documentclass[10pt]{sigplanconf}
\usepackage{etex}
\usepackage{float}
\usepackage{mathrsfs}
\usepackage{booktabs}
\usepackage{boxedminipage}
\usepackage[T1]{fontenc}
\usepackage[pdf]{pstricks}
\usepackage{epsfig}
\usepackage{multirow}
\usepackage{url}
\usepackage[normalem]{ulem}
\usepackage{schemepgm}
\usepackage{graphicx}
\usepackage{graphpap}
\usepackage{tabularx}
\usepackage{algorithmic}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{pstricks}
\usepackage{pst-text}
\usepackage{pst-node}
\usepackage{pst-tree}
\usepackage{pst-rel-points}
\usepackage{bcprules}
\input{fh_defs}
\def\drawplusplus#1#2#3{\hbox to 0pt{\hbox to #1{\hfill\vrule height #3 depth
      0pt width #2\hfill\vrule height #3 depth 0pt width #2\hfill
      }}\vbox to #3{\vfill\hrule height #2 depth 0pt width
      #1 \vfill}}
      %Poor man's typography
\def\concat{\mathrel{\drawplusplus {12pt}{0.4pt}{5pt}}}
      %It would be better to specify these in font-relative measures, but it 
      %probably doesn't scale anyway.
%\newcommand{\comment}[1]{}
\definecolor{Myblue}{rgb}{0.0,0.0,0.5}
\newcommand{\comment}[1]{{\color{Myblue}{\medskip \hrule\medskip
    #1 \medskip \hrule \medskip}}}
\newcommand{\alan}[1]{{\color{red}{\medskip \hrule\medskip
    #1 \medskip \hrule \medskip}}}
\newcommand{\blankout}[1]{}

@@

\begin{figure*}[t!]
\begin{center}
\begin{tabular}{|c|c|c|}
\hline
Premise & Transition & Rule name \\
\hline
\hline
          & $\rho, (\rho', x, e)\!:\!S, H, \kappa
  \rightsquigarrow \rho', S, H[\rho'(x) \mapsto \kappa], e$    &  \sc{const} \\
\hline
          & \shortstack[c]{$\rho, (\rho', z, e)\!:\!S, H, (\CONS~x~y)  \rightsquigarrow
$ \\  $\rho', S, H[z \mapsto (H(\rho(x)),H(\rho(y)))], e$}     &  \sc{cons} \\
\hline
$H(\rho(x)) \mbox{ is } (d_1, d_2)$ & $\rho, (\rho', z, e)\!:\!S, H,
(\CAR~x)  \rightsquigarrow \rho', S, H[\rho(z) \mapsto d_1], e$      &  \sc{car-whnf} \\
\hline
$H(\rho(x)) \mbox{ is } (e, \rho')$ & $\rho, S, H, (\CAR~x)  \rightsquigarrow
\rho'\oplus\rho, (\rho, x, (\CAR~x))\!:\!S, H, e$      &  \sc{car-clo} \\
\hline
$H(\rho(x)), H(\rho(y)) \in \mathbb{N}$
 & \shortstack[c]{$\rho, (\rho', z, e)\!:\!S, H, (+~x~y)  \rightsquigarrow$ \\
$\rho', S, H[\rho(z) \mapsto H(\rho(x)) + H(\rho(y))], e$}      &  \sc{prim-whnf} \\
\hline
$H(\rho(x)) \mbox{ is } (s, \rho')$ & $\rho, S, H, (+~x~y)  \rightsquigarrow
\rho'\oplus\rho, (\rho, x, (+~x~y))\!:\!S, H, e$      &  \sc{prim-1-clo} \\
\hline
$H(\rho(y)) \mbox{ is } (s, \rho')$ & $\rho, S, H, (+~x~y)  \rightsquigarrow
\rho'\oplus\rho, (\rho, x, (+~x~y))\!:\!S, H, e$      &  \sc{prim-2-clo} \\
\hline
\shortstack[c]{$\mathit{f}~\mbox{defined as}$ \\ $~(\DEFINE~(f~\myvec{y})~e_{\mathit{f}})$}  & $\rho, S, H, (f~\myvec{x})  \rightsquigarrow
[\myvec{y} \mapsto \rho(\myvec{x})], S, H, e_{\mathit{f}}$      &  \sc{funcall} \\
\hline
$\ell$ is a new location& \shortstack[c]{$\rho, S, H, (\LET~x\leftarrow s~\IN~e)
  \rightsquigarrow$ \\
$[x \mapsto \ell]\oplus\rho, S, H[\ell \mapsto (s, (x \mapsto
  \ell)\oplus\lfloor \rho\rfloor_{FV(S)\setminus\{x\}})], e$} & \sc{let} \\ 
\hline
$H(\rho(x)) \ne 0$ & $\rho, S, H, (\SIF~x~e_1~e_2)   \rightsquigarrow
\rho, S, H,  e_1$ & \sc{if-true} \\ 
\hline
$H(\rho(x)) = 0$ & $\rho, S, H, (\SIF~x~e_1~e_2)   \rightsquigarrow
\rho, S, H,  e_2$ & \sc{if-false} \\ 
\hline
$H(\rho(x)) = (e,\rho')$ & \shortstack[c]{$\rho, S, H,
  (\SIF~x~e_1~e_2)   \rightsquigarrow$ \\ 
$\rho'\oplus\rho, (\rho, x, (\IF~x~e_1~e_2))\!:\!S, H,  e$} & \sc{if-clo} \\ 
\hline
\shortstack[c]{$H(\rho(x))~\mbox{is}$ \\$\mbox{whnf with value}~v$}& $\rho, (\rho', z, e)\!:\!S, H,
(\RETURN~x)  \rightsquigarrow \rho', S, H[\rho(z) \mapsto v], e$ & \sc{return-whnf}\\   
\hline
$H(\rho(x)) = (e,\rho')$ & \shortstack[c]{$\rho, S, H, (\RETURN~x)
  \rightsquigarrow$ \\
$\rho'\oplus\rho,~ (\rho, x, (\RETURN~x))\!:\!S, H,  e$} & \sc{return-coo} \\
\hline
\end{tabular}
\caption{A small-step semantics for the language}\label{fig:lang-semantics}
\end{center}
\end{figure*}

@@

\pagebreak

@@

A variable is {\em live} it  there is a possibility of its value being
used  in future computation  and dead  if it  is definitely  not used.
Heap-allocated data needs a richer model of liveness which talks about
the  liveness of pointers  including fields  of $\CONS$  cells.  Using
$\acar$, $\acdr$  to represent access using $\CAR$  and $\CDR$ fields,
the  liveness  of the  structure  reachable  from  a variable  can  be
represented by a set of {\em access paths} i.e.  prefix-closed strings
from $\{\acar,\acdr\}^\ast$.   As an  example, if $x$  is a  list with
liveness      $\{\epsilon,     \acdr,      \acdr\acar,     \acdr\acdr,
\acdr\acdr\acar\}$, then future computations  can only refer up to the
second  and  third  members  of  $x$. In  this  notation,  $x  \mapsto
\{\epsilon\}$ represents access using  $x$ itself and $x \mapsto \{\}$
indicates that $x$ is dead.  A {\em liveness environment} is a mapping
from  variables  to   subsets  of  $\{\acar,\acdr\}^\ast$,  but  often
expressed as  a set, for example by  writing $\{x.\acar\acdr, x.\acdr,
y.\epsilon\}$               instead               of               $[x
  \mapsto\{\epsilon,\acar,\acar\acdr,\acdr\},     y\mapsto\{\epsilon\},
  z\mapsto\{\}]$.   We  associate  a  liveness environment  with  
program  points $\pi$.  The liveness  at $\pi:e$  is the  liveness just
before executing $e$.

@@

We use $\sigma$  to range over demands, $\alpha$  to range over access
paths  and $\Lv$ to  range over  liveness environments.   The notation
$\sigma_1\sigma_2$  denotes  the  set $\lbrace  \alpha_1\alpha_2  \mid
\alpha_1 \in \sigma_1, \alpha_2  \in \sigma_2\rbrace$.  Often we shall
abuse notation to  juxtapose an edge label and a  set of access paths:
$\acar\sigma$ is a shorthand for $\lbrace\acar\rbrace\sigma$.
%% Finally,  we use $\Lfonly$  to range  over {\em  demand transformers}.
%% These transform demands on a function call into demands for its formal
%% parameters:    given    a    user    function    $f$,    defined    by
%% $(\DEFINE\  (f\ x_1\  \ldots\ x_n)\  \  e_f)$ and  called with  demand
%% $\sigma$, then $\Lf{f}{i}{\sigma}$ is the liveness of $x_i$ at $e_f$.

@@

Figure~\ref{fig:live-judge}  describes  our  analysis.   Firstly,  the
function $\mathit{ref}$,  when given an  application $s$ and  a demand
$\sigma$,  returns the liveness  generated for  the free  variables of
$s$.   In  each  case,  the  null demand  ($\emptyset$)  generates  no
liveness.   Notice  that this  is  different  from  liveness of  eager
languages, where, for instance,  (\CDR~$x$) would be evaluated even if
the   demand   on   it   is  $\emptyset$   generating   the   liveness
$\{x.\epsilon\}$.  The  cases (\PRIM~x~y)  and (\NULLQ~x) are  easy. A
non-null  demand of  $\sigma$  on (\CDR~$x$),  is  transformed to  the
liveness  $x.\epsilon,  x.\acdr\sigma$.   In  an opposite  sense,  the
demand  of  $\acar\sigma$ on  (\CONS~$x$~$y$)  is  transformed to  the
demand  $\sigma$  on $x$.   Note  that  \CONS\  does not,  by  itself,
dereference its arguments. Constants generate no liveness.

For  the  case  of a  user-function  call  we  use a  third  parameter
$\Lfonly$  that  represents the  summaries  of  all  functions in  the
program. $\Lfonly$ expresses how the  demand $\sigma$ on the result of
a function call is transformed  into liveness on its parameters at the
beginning of the call.  $\Lfonly$  is determined by the judgement form
$\mathit{Prog}    \len   \Lfonly$    using   inference    rule   ({\sc
  live-define}). This states the  fixpoint property to be satisfied by
$\Lfonly$, namely, the demand transformation assumed for each function
in  the  program should  be  the  same  as the  demand  transformation
calculated from their bodies. The algorithmic formulation of this rule
through  a  grammar will  be  described  in [???].   This yields  the  least
solution  satisfying  the  fixpoint  property  and  thus  ensures  the
greatest amount of garbage collected.

%%   Analogously   to  classical  liveness  being
%% computed  as  a  solution  of  dataflow  equations,  we  require,  via
%% inference   rule  ({\sc  live-define}),   $\Lfonly$  to   satisfy  the
%% fixed-point property that:  when we assume $\Lfonly$ to  be the family
%% of demand transformers for the program then the calculated liveness of
%% each function  body $\mathcal{L}(e_f,\sigma,\Lfonly)$ agrees  with the
%% assumed $\Lfonly$.   As usual, there  are often multiple  solutions to
%% $\Lfonly$;  all are  safe (see  Section~\ref{sec:correctness})  but we
%% prefer  the  least  one  as  giving  the  least  liveness  subject  to
%% safety---and hence greatest amount of garbage collected.

We now describe the  propagation of liveness through expressions. This
brings  us to  the  second way  in  which liveness  analysis for  lazy
languages  differ from  eager languages---the point where  an
expression gets evaluated cannot be statically determined.  To see how
this  is handled,  consider  the liveness  analysis  of the  following
function:
%%%%%%%%%%%%%%%%%%%%%%

@@

%% To  deal  with the  first  issue, we  apply  the  results of  liveness
%% analysis only  to fully evaluated  data.  If the marking  phase during
%% garbage collection leads to a closure, and the closure is not dead, we
%% mark the entire reachable memory  under the closure for collection.  A
%% better scheme  would be  to garbage collect  the closure based  on the
%% liveness  at the point  of its  creation.  This  is sound  because our
%% language is statically scoped.  However, this is work in progress, and
%% the method described in this paper switches over to reachability while
%% collecting closures.

@@

Assume that the calls to the $\pf$ place an overall demand of $\sigma$
on the body  of \pf.  For obvious reasons, this is  also the demand on
the return  values $u$ and $y$,  and thus the liveness  at the program
points  $\pi_7$  and  $\pi_8$  are $\lbrace  \pu.\sigma  \rbrace$  and
$\{\py.\sigma\}$.  The liveness at  $\pi_7$ and $\pi_8$ are propagated
backwards as in traditional liveness analysis.

Since   \pu\  is  $(\CDR~   \pw)$,  the   liveness  of   $\sigma$  for
\pu\ translates  into a demand  of $\sigma$ on $(\CDR~\pw)$,  which in
turn generates  a liveness of $\{\pw,~\pw.1\sigma\}$  at $\pi_6$.  The
liveness of  \pu\ is  also killed at  $\pi_6$ since  it is not  in the
scope of \pu.   However, for simplicity of exposition,  we do not kill
livenesses of  variables in our formulation.  This  does not introduce
any imprecision  since \pu\ is yet  to come into  existence at $\pi_6$
and its  liveness will  not be consulted  by the garbage  collector at
this point.
%==============================================================
\begin{figure*}[t]  
\begin{eqnarray*}
\mathit{ref\/}(\kappa,\sigma,\Lfonly)
          &=& \{\,\} \mbox{, for $\kappa$ a constant, including $\NIL$}\\
\mathit{ref\/}((\CONS~x~y),\sigma,\Lfonly)
          &=& \{x.\alpha \mid \acar\alpha \in \sigma\} \cup \{y.\alpha \mid \acdr\alpha \in \sigma\} \\
\mathit{ref\/}((\CAR~x),\sigma,\Lfonly)
          &=&    \begin{array}{l l}
                    \{x.\epsilon\} \cup \{x.\acar\alpha \mid \alpha \in
\sigma\}, & \mbox{if}~\sigma \ne \emptyset\\
                    \emptyset  & \mbox{otherwise}
                 \end{array} \\
\mathit{ref\/}((\CDR~x),\sigma,\Lfonly)
          &=&    \begin{array}{l l}
                    \{x.\epsilon\} \cup \{x.\acdr\alpha \mid \alpha \in
\sigma\}, & \mbox{if}~\sigma \ne \emptyset\\
                    \emptyset  & \mbox{otherwise}
                 \end{array} \\
\mathit{ref\/}((\PRIM~x~y),\sigma,\Lfonly)
          &=&    \begin{array}{l l}
                    \{x.\epsilon, y.\epsilon\},  & \mbox{if}~\sigma \ne \emptyset\\
                    \emptyset  & \mbox{otherwise}
                 \end{array} \\
\mathit{ref\/}((\NULLQ~x),\sigma,\Lfonly)
          &=&    \begin{array}{l l}
                    \{x.\epsilon\},  & \mbox{if}~\sigma \ne \emptyset\\
                    \emptyset  & \mbox{otherwise}
                 \end{array} \\
\mathit{ref\/}((f~\myvec{x}),\sigma,\Lfonly)
%          &=& \bigcup_{i=1}^n y_i.\Lf{f}{i}{\sigma}
          &=&  \begin{array}{@{}l}  % to discourage \displaystyle
               \bigcup_{i=1}^n x_i.\Lf{f}{i}{\sigma}
               \end{array}
%          &=& \bigcup \{y_i.\Lf{f}{i}{\sigma} \mid i=1,\ldots, n\}
\\[1ex]
\mathcal{L}((\RETURN~x),\sigma,\Lfonly) &=& \{l \mapsto x.\sigma\}, \makebox[0mm]{\hspace*{3cm} where $l$ is a new label} \\
\mathcal{L}((\SIF~x~e_1~e_2),\sigma,\Lfonly) &=&
        \begin{array}{l l}
                    \mathcal{L}(e_1,\sigma,\Lfonly) \uplus
        \mathcal{L}(e_2,\sigma,\Lfonly) \uplus
        \{l \mapsto  \{x.\epsilon\}\},  & \mbox{if}~\sigma \ne \emptyset\\
        \emptyset_{\cal{M}?}  & \mbox{otherwise}
                 \end{array} \\
 \makebox[0mm]{\hspace*{3.8cm} where $l$ is a new label}\\
\mathcal{L}(\LET~x \leftarrow~s~\IN~e),\sigma,\Lfonly) &=&
        \{ l \mapsto \Ldonly_{x \leftarrow s}(\Lv(x)) \mid l \mapsto \Lv
        \in \mathcal{M}\}
\mbox{ where } \mathcal{M} = \mathcal{L}(e,\sigma,\Lfonly)\\
\Ldonly_{x \leftarrow s} &=& \Lfonly_f, \mbox{where}~f~\mbox{is a new function}\\
 \makebox[0mm]{\hspace*{7cm}   
 $(\DEFINE~(f\,x_1\,x_2\, \ldots \, x_n)~(\SIF\,*~x~s[(f\,x_1\,
           x_2\, \ldots\, x_n)/x]))$,} \\
 \makebox[0mm]{\hspace*{5.5cm} and 
     $(\ x_1\, \ldots\, x_n)$ are the variables in $s$}
\end{eqnarray*}
\begin{minipage}{0.85\textwidth}
\infrule[live-define]
        {\mathcal{L}(e_f,\sigma,\Lfonly) = 
           \bigcup_{i=1}^n z_i.\Lf{f}{i}{\sigma}
              \mbox{ for each $f$ and $\sigma$}
        }
        { d_1 \ldots d_k \len \Lfonly
\\ \makebox[0mm]{where
     $(\DEFINE\ (f\ z_1\ \ldots\ z_n)\ \ e_f)$ is a member of $d_1 \ldots d_k$}}
\end{minipage}
  \caption{Liveness equations and judgement rule}\label{fig:live-judge}
\end{figure*}
%======================================================================================

>>>>

\documentclass[10pt]{sigplanconf}
\usepackage{etex}
\usepackage{float}
\usepackage{mathrsfs}
\usepackage{booktabs}
\usepackage{boxedminipage}
\usepackage[T1]{fontenc}
\usepackage[pdf]{pstricks}
\usepackage{epsfig}
\usepackage{multirow}
\usepackage{url}
\usepackage[normalem]{ulem}
\usepackage{schemepgm}
\usepackage{graphicx}
\usepackage{graphpap}
\usepackage{tabularx}
\usepackage {algorithmic}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{pstricks}
\usepackage{pst-text}
\usepackage{pst-node}
\usepackage{pst-tree}
\usepackage{pst-rel-points}
\usepackage{bcprules}
%\usepackage{program}
\input{fh_defs}
\def\drawplusplus#1#2#3{\hbox to 0pt{\hbox to #1{\hfill\vrule height #3 depth
      0pt width #2\hfill\vrule height #3 depth 0pt width #2\hfill
      }}\vbox to #3{\vfill\hrule height #2 depth 0pt width
      #1 \vfill}}
      %Poor man's typography
\def\concat{\mathrel{\drawplusplus {12pt}{0.4pt}{5pt}}}
      %It would be better to specify these in font-relative measures, but it 
      %probably doesn't scale anyway.
\newcommand{\mycomment}[1]{}
\definecolor{Myblue}{rgb}{0.0,0.0,0.5}
\newcommand{\comment}[1]{{\color{Myblue}{\medskip \hrule\medskip
    #1 \medskip \hrule \medskip}}}
\newcommand{\alan}[1]{{\color{red}{\medskip \hrule\medskip
    #1 \medskip \hrule \medskip}}}
\newcommand{\blankout}[1]{}

@@

\begin{figure*}[t!]
\begin{center}
\begin{tabular}{|c|c|c|}
\hline
Premise & Transition & Rule name \\
\hline
\hline
          & $\rho, (\rho', x, e)\!:\!S, H, \kappa
  \rightsquigarrow \rho', S, H[\rho'(x) \mapsto \kappa], e$    &  \sc{const} \\
\hline
          & \shortstack[c]{$\rho, (\rho', z, e)\!:\!S, H, (\CONS~x~y)  \rightsquigarrow
$ \\  $\rho', S, H[z \mapsto (H(\rho(x)),H(\rho(y)))], e$}     &  \sc{cons} \\
\hline
$H(\rho(x)) \mbox{ is } (d_1, d_2)$ & $\rho, (\rho', z, e)\!:\!S, H,
(\CAR~x)  \rightsquigarrow \rho', S, H[\rho(z) \mapsto d_1], e$      &  \sc{car-whnf} \\
\hline
$H(\rho(x)) \mbox{ is } (e, \rho')$ & $\rho, S, H, (\CAR~x)  \rightsquigarrow
\rho'\oplus\rho, (\rho, x, (\CAR~x))\!:\!S, H, e$      &  \sc{car-clo} \\
\hline
$H(\rho(x)), H(\rho(y)) \in \mathbb{N}$
 & \shortstack[c]{$\rho, (\rho', z, e)\!:\!S, H, (+~x~y)  \rightsquigarrow$ \\
$\rho', S, H[\rho(z) \mapsto H(\rho(x)) + H(\rho(y))], e$}      &  \sc{prim-whnf} \\
\hline
$H(\rho(x)) \mbox{ is } (s, \rho')$ & $\rho, S, H, (+~x~y)  \rightsquigarrow
\rho'\oplus\rho, (\rho, x, (+~x~y))\!:\!S, H, e$      &  \sc{prim-1-clo} \\
\hline
$H(\rho(y)) \mbox{ is } (s, \rho')$ & $\rho, S, H, (+~x~y)  \rightsquigarrow
\rho'\oplus\rho, (\rho, x, (+~x~y))\!:\!S, H, e$      &  \sc{prim-2-clo} \\
\hline
\shortstack[c]{$\mathit{f}~\mbox{defined as}$ \\ $~(\DEFINE~(f~\myvec{y})~e_{\mathit{f}})$}  & $\rho, S, H, (f~\myvec{x})  \rightsquigarrow
[\myvec{y} \mapsto \rho(\myvec{x})], S, H, e_{\mathit{f}}$      &  \sc{funcall} \\
\hline
$\ell$ is a new location& \shortstack[c]{$\rho, S, H, (\LET~x\leftarrow s~\IN~e)
  \rightsquigarrow$ \\
$[x \mapsto \ell]\oplus\rho, S, H[\ell \mapsto (s, (x \mapsto
  \ell)\oplus\lfloor \rho\rfloor_{FV(S)\setminus\{x\}})], e$} & \sc{let} \\ 
\hline
$H(\rho(x)) \ne 0$ & $\rho, S, H, (\SIF~x~e_1~e_2)   \rightsquigarrow
\rho, S, H,  e_1$ & \sc{if-true} \\ 
\hline
$H(\rho(x)) = 0$ & $\rho, S, H, (\SIF~x~e_1~e_2)   \rightsquigarrow
\rho, S, H,  e_2$ & \sc{if-false} \\ 
\hline
$H(\rho(x)) = (e,\rho')$ & \shortstack[c]{$\rho, S, H,
  (\SIF~x~e_1~e_2)   \rightsquigarrow$ \\ 
$\rho'\oplus\rho, (\rho, x, (\SIF~x~e_1~e_2))\!:\!S, H,  e$} & \sc{if-clo} \\ 
\hline
\shortstack[c]{$H(\rho(x))~\mbox{is}$ \\$\mbox{whnf with value}~v$}& $\rho, (\rho', z, e)\!:\!S, H,
(\RETURN~x)  \rightsquigarrow \rho', S, H[\rho(z) \mapsto v], e$ & \sc{return-whnf}\\   
\hline
$H(\rho(x)) = (e,\rho')$ & \shortstack[c]{$\rho, S, H, (\RETURN~x)
  \rightsquigarrow$ \\
$\rho'\oplus\rho,~ (\rho, x, (\RETURN~x))\!:\!S, H,  e$} & \sc{return-coo} \\
\hline
\end{tabular}
\caption{A small-step semantics for the language}\label{fig:lang-semantics}
\end{center}
\end{figure*}

@@

@@

A variable is {\em live} if  there is a possibility of its value being
used  in future computation  and dead  if it  is definitely  not used.
Heap-allocated data needs a richer model of liveness which talks about
liveness  of  pointers  including  fields  of  $\CONS$  cells.   Using
$\acar$, $\acdr$  to represent access using $\CAR$  and $\CDR$ fields,
the  liveness  of the  structure  reachable  from  a variable  can  be
represented by a set of {\em access paths} i.e.  prefix-closed strings
from $\{\acar,\acdr\}^\ast$.   As an  example, if $x$  is a  list with
liveness      $\{\epsilon,     \acdr,      \acdr\acar,     \acdr\acdr,
\acdr\acdr\acar\}$, then future computations  can only refer up to the
second and  third members of $x$.   A {\em liveness  environment} is a
mapping from variables to subsets of $\{\acar,\acdr\}^\ast$, but often
expressed as  a set, for example by  writing $\{x.\epsilon, x.\acdr, x.\acdr\acdr, 
y.\epsilon\}$               instead               of               $[x
  \mapsto\{\epsilon,\acdr,\acdr\acdr\},     y\mapsto\{\epsilon\},
  z\mapsto\{\}]$.   In  this   notation,   $x  \mapsto   \{\epsilon\}$
represents access using $x$ itself and $x \mapsto \{\}$ indicates that
$x$ is dead.  We associate  a liveness environment with program points
$\pi$.  The liveness at $\pi:e$  is the liveness just before executing
$e$.

@@

We use $\sigma$  to range over demands, $\alpha$  to range over access
paths  and $\Lv$ to  range over  liveness environments.   The notation
$\sigma_1\sigma_2$  denotes  the  set $\lbrace  \alpha_1\alpha_2  \mid
\alpha_1 \in \sigma_1, \alpha_2  \in \sigma_2\rbrace$.  Often we shall
abuse notation to  juxtapose an edge label and a  set of access paths:
$\acar\sigma$ is a shorthand for $\lbrace\acar\rbrace\sigma$. A more
elaborate explanation of these notions appear in \cite{asati14lgc}

%% Finally, we use $\Lfonly$  to range over {\em demand transformers}.
%% These transform  demands on  a function call  into demands  for its
%% formal  parameters:   given  a   user  function  $f$,   defined  by
%% $(\DEFINE\ (f\  x_1\ \ldots\ x_n)\  \ e_f)$ and called  with demand
%% $\sigma$,  then $\Lf{f}{i}{\sigma}$  is  the liveness  of $x_i$  at
%% $e_f$.

@@

Figure~\ref{fig:live-judge}  describes  our  analysis  which  has  two
parts. The function $\mathit{ref}$,  when given an application $s$ and
a demand $\sigma$, returns  the incremental liveness generated for the
free  variables   of  $s$  due  to  the   application.   The  function
$\mathcal{L}$  uses   $\mathit{ref}$  to  propagate   liveness  across
expressions.

 In each  rule defining $\mathit{ref}$, the  null demand ($\emptyset$)
 generates no  liveness.  Notice that this is  different from liveness
 of  eager  languages,  where,   for  instance,  (\CDR~$x$)  would  be
 evaluated even for the null  demand, and this would require an access
 using  $x$.   A  non-null   demand  of  $\sigma$  on  (\CDR~$x$),  is
 transformed  to  the  liveness  $\{x.\epsilon, x.\acdr\sigma\}$.   In  an
 opposite  sense, the  demand of  $\acdr\sigma$ on  (\CONS~$y$~$z$) is
 transformed to  the demand  $\sigma$ on $z$.   Note that  \CONS\ does
 not, by itself, dereference  its arguments. The rules for (\PRIM~x~y)
 and (\NULLQ~x) are on similar lines. Constants generate no liveness.

For  the  case  of a  user-function  call  we  use a  third  parameter
$\Lfonly$  that  represents the  summaries  of  all  functions in  the
program. $\Lfonly_f$ (the component of $\Lfonly$ for a specific
function $f$),  expresses how the  demand $\sigma$ on the result of
a call to $f$ is transformed  into the liveness of its parameters at the
beginning of the call.  $\Lfonly$  is determined by the judgement form
$\mathit{Prog} \len \Lfonly$ using inference rule ({\sc live-define}).
This  rule  describes the  fixed-point  property  to  be satisfied  by
$\Lfonly$, namely, the demand transformation assumed for each function
in  the  program should  be  the  same  as the  demand  transformation
calculated    from    their   bodies.    As    we    shall   see    in
Section~\ref{sec:grammar-formulation},  the  least solution  satisfying
the  rule  is obtained  by  converting  the  rule into  an  equivalent
grammar.  The least solution also  ensures that the greatest amount of
garbage is collected.

%%   Analogously to classical liveness being computed as a solution of
%% dataflow   equations,  we   require,  via   inference   rule  ({\sc
%% live-define}), $\Lfonly$ to  satisfy the fixed-point property that:
%% when we  assume $\Lfonly$ to  be the family of  demand transformers
%% for the program then the  calculated liveness of each function body
%% $\mathcal{L}(e_f,\sigma,\Lfonly)$    agrees   with    the   assumed
%% $\Lfonly$.   As  usual,  there  are  often  multiple  solutions  to
%% $\Lfonly$; all are  safe (see Section~\ref{sec:correctness}) but we
%% prefer  the least  one  as  giving the  least  liveness subject  to
%% safety---and hence greatest amount of garbage collected.

We next  describe the function $\mathcal{L}$  that propagates liveness
across  expressions.   This brings  us  to  the  second way  in  which
liveness analysis for lazy languages differ from eager languages---the
point  where  an  expression   gets  evaluated  cannot  be  statically
determined.  To  deal with this, we introduce  flow insensitivity into
our  analysis.

%%%%%%%%%%%%%%%%%%%%%%

@@

%% To  deal with the  first issue,  we apply  the results  of liveness
%% analysis only to fully evaluated data.  If the marking phase during
%% garbage collection leads to a closure, and the closure is not dead,
%% we  mark  the  entire   reachable  memory  under  the  closure  for
%% collection.   A  better scheme  would  be  to  garbage collect  the
%% closure based on  the liveness at the point  of its creation.  This
%% is sound because our  language is statically scoped.  However, this
%% is  work  in progress,  and  the  method  described in  this  paper
%% switches over to reachability while collecting closures.

@@

Consider  an analysis of  the body  of $\pf$  with a  demand $\sigma$.
Clearly, this is also the demand on the return values $u$ and $y$
giving the liveness  at  the  program points  $\pi_7$  and $\pi_8$  as
$\lbrace  \pu.\sigma \rbrace$ and  $\{\py.\sigma\}$.  The  liveness at
$\pi_7$  and  $\pi_8$  are  propagated  backwards  as  in  traditional
liveness analysis.

Since   \pu\  is  $(\CDR~   \pw)$,  the   liveness  of   $\sigma$  for
\pu\ translates  into a demand  of $\sigma$ on $(\CDR~\pw)$,  which in
turn  generates  a  liveness  of $\{\pw.\epsilon,  ~\pw.1\sigma\}$  at
$\pi_7$.  The liveness of \pu\ should be killed at $\pi_7$ since it is
not in the scope of \pu.  However, this does not cause any imprecision
in the way  of the GC marking more  cells as live, since \pu\  is yet to
come into existence at $\pi_7$  and its liveness will not be consulted
by the garbage collector at this point.
%==============================================================
\begin{figure*}[t]  
\begin{eqnarray*}
\mathit{ref\/}(\kappa,\sigma,\Lfonly)
          &=& \{\,\} \mbox{, for $\kappa$ a constant, including $\NIL$}\\
\mathit{ref\/}((\CONS~x~y),\sigma,\Lfonly)
          &=& \{x.\alpha \mid \acar\alpha \in \sigma\} \cup \{y.\alpha \mid \acdr\alpha \in \sigma\} \\
\mathit{ref\/}((\CAR~x),\sigma,\Lfonly)
          &=&    \begin{array}{l l}
                    \{x.\epsilon\} \cup \{x.\acar\alpha \mid \alpha \in
\sigma\}, & \mbox{if}~\sigma \ne \emptyset\\
                    \emptyset  & \mbox{otherwise}
                 \end{array} \\
\mathit{ref\/}((\CDR~x),\sigma,\Lfonly)
          &=&    \begin{array}{l l}
                    \{x.\epsilon\} \cup \{x.\acdr\alpha \mid \alpha \in
\sigma\}, & \mbox{if}~\sigma \ne \emptyset\\
                    \emptyset  & \mbox{otherwise}
                 \end{array} \\
\mathit{ref\/}((\PRIM~x~y),\sigma,\Lfonly)
          &=&    \begin{array}{l l}
                    \{x.\epsilon, y.\epsilon\},  & \mbox{if}~\sigma \ne \emptyset\\
                    \emptyset  & \mbox{otherwise}
                 \end{array} \\
\mathit{ref\/}((\NULLQ~x),\sigma,\Lfonly)
          &=&    \begin{array}{l l}
                    \{x.\epsilon\},  & \mbox{if}~\sigma \ne \emptyset\\
                    \emptyset  & \mbox{otherwise}
                 \end{array} \\
\mathit{ref\/}((f~\myvec{x}),\sigma,\Lfonly)
%          &=& \bigcup_{i=1}^n y_i.\Lf{f}{i}{\sigma}
          &=&  \begin{array}{@{}l}  % to discourage \displaystyle
               \bigcup_{i=1}^n x_i.\Lf{f}{i}{\sigma}
               \end{array}
%          &=& \bigcup \{y_i.\Lf{f}{i}{\sigma} \mid i=1,\ldots, n\}
\\[1ex]
\mathcal{L}((\RETURN~x),\sigma,\Lfonly) &=& \{l \mapsto x.\sigma\}, \makebox[0mm]{\hspace*{3cm} where $l$ is a new label} \\
\mathcal{L}((\SIF~x~e_1~e_2),\sigma,\Lfonly) &=&
        \begin{array}{l l}
                    \mathcal{L}(e_1,\sigma,\Lfonly) \uplus
        \mathcal{L}(e_2,\sigma,\Lfonly) \uplus
        \{l \mapsto  \{x.\epsilon\}\},  & \mbox{if}~\sigma \ne \emptyset\\
        \emptyset_{\cal{M}?}  & \mbox{otherwise}
                 \end{array} \\
 \makebox[0mm]{\hspace*{3.8cm} where $l$ is a new label}\\
\mathcal{L}(\LET~x \leftarrow~s~\IN~e),\sigma,\Lfonly) &=&
        \{ l \mapsto \Ldonly_{x \leftarrow s}(\Lv(x)) \mid l \mapsto \Lv
        \in \mathcal{M}\}
\mbox{ where } \mathcal{M} = \mathcal{L}(e,\sigma,\Lfonly)\\
\Ldonly_{x \leftarrow s} &=& \Lfonly_f, \mbox{where}~f~\mbox{is a new function}\\
 \makebox[0mm]{\hspace*{7cm}   
 $(\DEFINE~(f\,x_1\,x_2\, \ldots \, x_n)~(\SIF\,*~x~s[(f\,x_1\,
           x_2\, \ldots\, x_n)/x]))$,} \\
 \makebox[0mm]{\hspace*{5.5cm} and 
     $(\ x_1\, \ldots\, x_n)$ are the variables in $s$}
\end{eqnarray*}
\begin{minipage}{0.85\textwidth}
\infrule[live-define]
        {\mathcal{L}(e_f,\sigma,\Lfonly) = 
           \bigcup_{i=1}^n z_i.\Lf{f}{i}{\sigma}
              \mbox{ for each $f$ and $\sigma$}
        }
        { d_1 \ldots d_k \len \Lfonly
\\ \makebox[0mm]{where
     $(\DEFINE\ (f\ z_1\ \ldots\ z_n)\ \ e_f)$ is a member of $d_1 \ldots d_k$}}
\end{minipage}
  \caption{Liveness equations and judgement rule}\label{fig:live-judge}
\end{figure*}
%======================================================================================
